---
title: "Time Series Project"
output: word_document
date: "2023-08-17"
author: catherine
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
```
#installing packages

```{r}
install.packages("tidyverse")   # For data manipulation
install.packages("ggplot2")     # For data visualization
install.packages("forecast")    # For time series forecasting and ARIMA
install.packages("tseries")     #for time series



```


#Packages used
```{r}
library(tidyverse)   # For data manipulation
library(ggplot2)     # For data visualization
library(forecast)    # For time series forecasting and ARIMA
library(tseries)


```



# 1. Problem Definition:***************************



# 2. Data Collection:


# 3. Data Exploration and Preprocessing:
```{r}

# Loading data

data <- read.csv("data/data.csv")

View(data)
str(data)

# Cheking for correct data types

#Check for outliers

# Get summary statistics of the data.***
summary(data)

clean_data <-  na.omit(data)

summary(clean_data)

# plotting a time series plot to visualize how the cases of Leishmaniasis vary over time
ggplot(clean_data, aes(x = Year, y = Reported.cases.Leishmaniasis)) +
  geom_line() +
  labs(title = "Cases of Leishmaniasis Over Time", x = "Year", y = "Cases")

```


# 4. Stationarity Check:
```{r}

adf_test <- adf.test(clean_data$Reported.cases.Leishmaniasis)
adf_test
# using Augmented Dickey-Fuller test for this If the p-value is less than a significance level (e.g., 0.05), you can reject the null hypothesis and consider the series stationary.

```

# 5. Differencing:

```{r}

#If your data is not stationary, apply differencing to make it stationary.
data$diff_cases <- diff(data$cases_of_Leishmaniasis)


```

# 6. Autocorrelation and Partial Autocorrelation (ACF and PACF) Analysis:
```{r}
# Assuming "data" is your data frame and "cases_of_Leishmaniasis" is the column of interest
acf_plot <- acf(data$cases_of_Leishmaniasis)
pacf_plot <- pacf(data$cases_of_Leishmaniasis)

# Plot ACF and PACF
plot(acf_plot)
plot(pacf_plot)

```

# 8. Model Fitting:

```{r}
#Model Selection
#Based on the ACF and PACF plots, choose the order of the ARIMA model (p, d, q).

# Example: ARIMA(1,1,1)
model <- arima(data$diff_cases, order = c(1, 1, 1))


```

# 9. Model Evaluation:
```{r}
# Calculate accuracy metrics
accuracy_metrics <- accuracy(forecasted_values, data$cases_of_Leishmaniasis)
accuracy_metrics

```

# 10. Forecasting:
```{r}
forecast_data <- forecast(model, h = 10)  # Forecasting 10 periods ahead
plot(forecast_data)

```

# 11. Model Testing:

```{r}
# Assuming "train_data" and "test_data" are your training and testing data subsets
model <- arima(train_data$cases_of_Leishmaniasis, order = c(p, d, q))
forecasted_values <- forecast(model, h = nrow(test_data))
accuracy_metrics <- accuracy(forecasted_values, test_data$cases_of_Leishmaniasis)
accuracy_metrics

```

# 12. Hyperparameter Tuning:
#Hyperparameter tuning involves finding the best values for the parameters of the ARIMA model that optimize its performance. #You can use techniques like grid search or optimization algorithms. Consider using libraries like forecast or caret for this #purpose.

# 13. Final Model Deployment (Optional):
# 14. Documentation and Reporting: