---
title: "Time Series Project"
output: word_document
date: "2023-08-17"
author: catherine Nyaguthii Mwai
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
```
#installing packages

```{r}
install.packages("tidyverse")   # For data manipulation
install.packages("ggplot2")     # For data visualization
install.packages("forecast")    # For time series forecasting and ARIMA
install.packages("tseries")     #for time series



```


#Packages used
```{r}
library(tidyverse)   # For data manipulation
library(ggplot2)     # For data visualization
library(forecast)    # For time series forecasting and ARIMA
library(tseries)


```



# 1. Problem Definition:***************************



# 2. Data Collection:


# 3. Data Exploration and Preprocessing:
```{r}

# Loading data

data <- read.csv("data/data.csv")

View(data)
str(data)

# Cheking for correct data types

#Check for outliers

# Get summary statistics of the data.***
summary(data)

clean_data <-  na.omit(data)

summary(clean_data)



```

```{r}

# Calculate the index corresponding to the first 70% of the data
data_length <- nrow(clean_data)
index_70_percent <- round(data_length * 0.7)

# Split the data into training and test sets
train_data <- clean_data[1:index_70_percent, ]
test_data <- clean_data[(index_70_percent + 1):data_length, ]

# Print the selected data
print("Training Data:")
print(train_data)

print("Test Data:")
print(test_data)









```



```{r}


# plotting a time series plot to visualize how the cases of Leishmaniasis vary over time
ggplot(train_data, aes(x = Year, y = Reported.cases.Leishmaniasis)) +
  geom_line() +
  labs(title = "Cases of Leishmaniasis Over Time", x = "Year", y = "Leishmaniasis Cases")

```

```


```{r}

adf_test <- adf.test(train_data$Reported.cases.Leishmaniasis)
adf_test
# using Augmented Dickey-Fuller test for this If the p-value is less than a significance level (e.g., 0.05), you can reject the null hypothesis and consider the series stationary.



```




# 4. Stationarity Check:
Null Hypothesis :The time series is non-stationary.

Alternative Hypothesis : The time series is stationary.




```{r}



adf_test <- adf.test(clean_data$Reported.cases.Leishmaniasis)
adf_test
# using Augmented Dickey-Fuller test for this If the p-value is less than a significance level (e.g., 0.05), you can reject the null hypothesis and consider the series stationary.

```

# 5. Differencing:

```{r}

#If your data is not stationary, apply differencing to make it stationary.
original_data <- clean_data$Reported.cases.Leishmaniasis
order_of_differencing <- 0
adf_p_value <- 1

while (adf_p_value >= 0.05) {  # Check for stationarity using ADF test
  if (order_of_differencing > 0) {
    differenced_data <- diff(original_data, differences = order_of_differencing)
  } else {
    differenced_data <- original_data
  }

  adf_result <- adf.test(differenced_data)
  adf_p_value <- adf_result$p.value
  
  if (adf_p_value >= 0.05) {
    order_of_differencing <- order_of_differencing + 1
    original_data <- differenced_data
  }
}

cat("Order of differencing:", order_of_differencing, "\n")
cat("Final ADF p-value:", adf_p_value, "\n")




```
Debugging

```{r}

# Assuming 'clean_data' is a data frame with a 'Year' column and the differenced data in 'differenced_data' column
Year <- clean_data$Year[-10]
View(Year)
View(original_data)

ggplot(clean_data, aes(x = Year, y = differenced_data)) +
  geom_line() +
  labs(title = "Differenced Cases of Leishmaniasis Over Time", x = "Year", y = "Differenced Cases")


```

```{r}

```



# 6. Autocorrelation and Partial Autocorrelation (ACF and PACF) Analysis:
```{r}
# Assuming "data" is your data frame and "cases_of_Leishmaniasis" is the column of interest
acf_plot <- acf(original_data)
pacf_plot <- pacf(original_data)

par(mfrow = c(1, 2))

# Plot ACF and PACF
plot(acf_plot)
plot(pacf_plot)

#According to this output, we will use ARIMA(1,4,0)
```

# 7. Data splitting to test and train data set
# 8. Model Fitting:

```{r}
#Model Selection
#Based on the ACF and PACF plots, choose the order of the ARIMA model (p, d, q).

# Example: ARIMA(1,1,1)
model <- arima(data$diff_cases, order = c(1, 1, 1))


```

# 9. Model Evaluation:
```{r}
# Calculate accuracy metrics
accuracy_metrics <- accuracy(forecasted_values, data$cases_of_Leishmaniasis)
accuracy_metrics

```

# 10. Forecasting:
```{r}
forecast_data <- forecast(model, h = 10)  # Forecasting 10 periods ahead
plot(forecast_data)

```

# 11. Model Testing:

```{r}
# Assuming "train_data" and "test_data" are your training and testing data subsets
model <- arima(train_data$cases_of_Leishmaniasis, order = c(p, d, q))
forecasted_values <- forecast(model, h = nrow(test_data))
accuracy_metrics <- accuracy(forecasted_values, test_data$cases_of_Leishmaniasis)
accuracy_metrics

```

# 12. Hyperparameter Tuning:
#Hyperparameter tuning involves finding the best values for the parameters of the ARIMA model that optimize its performance. #You can use techniques like grid search or optimization algorithms. Consider using libraries like forecast or caret for this #purpose.

# 13. Final Model Deployment (Optional):
# 14. Documentation and Reporting: